---
title: "Computational Stats Final Project"
author: "Daniel J. Park"
date: "3/18/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup 
```{r}
library(mgcv)
library(MASS)
library(stringr)
library(gamm4)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(viridis)
library(cowplot)
library(kableExtra)
library(docxtools)
library(knitr)
library(tibble)
library(dplyr)
library(gratia)
library(latex2exp)
library(GGally)
library(car)
library(tidyverse)
library(glmnet)

#Set the default theme for ggplot objects to theme_bw()
theme_set(theme_bw())
theme_update(panel.grid = element_blank())
```

# Code to reproduce CO2 analysis

```{r}
#The default CO2 plant variable is ordered;
#This recodes it to an unordered factor
CO2 <- transform(CO2, Plant_uo=factor(Plant, ordered=FALSE))

CO2_vis_plot <- ggplot(CO2, aes(x=conc, 
                                y=uptake, 
                                group=Plant,
                                color=Plant, 
                                lty=Plant)) +
  geom_point() +
  geom_line() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})), 
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))

CO2_vis_plot
```

## Model 1: Single global smoother
```{r}
CO2_modG <- gam(log(uptake) ~ s(log(conc), k=5, bs="tp") +
                  s(Plant_uo, k=12, bs="re"),
                data=CO2, method="REML", family="gaussian")
  
#plot the default gratia plot for the CO2 model
draw(CO2_modG)
```

```{r}
# setup prediction data
CO2_modG_pred <- with(CO2,
                      expand.grid(conc=seq(min(conc), max(conc), length=100),
                                  Plant_uo=levels(Plant_uo)))

# make the prediction, add this and a column of standard errors to the prediction
# data.frame. Predictions are on the log scale.
CO2_modG_pred <- cbind(CO2_modG_pred,
                       predict(CO2_modG, 
                               CO2_modG_pred, 
                               se.fit=TRUE, 
                               type="response"))

# make the plot. Note here the use of the exp() function to back-transform the
# predictions (which are for log-uptake) to the original scale
ggplot(data=CO2, aes(x=conc, y=uptake, color=Plant_uo, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(fit - 2*se.fit), ymax=exp(fit + 2*se.fit), x=conc),
              data=CO2_modG_pred, 
              alpha=0.3, 
              inherit.aes=FALSE) +
  geom_line(aes(y=exp(fit)), data=CO2_modG_pred) +
  geom_point() +
  labs(title="Model 1 Predictions",
       x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```

## Model 2: Global smoother plus group-level smoothers with similar wiggliness
```{r}
CO2_modGS <- gam(log(uptake) ~ s(log(conc), k=5, m=2) + 
                   s(log(conc), Plant_uo, k=5,  bs="fs", m=2),
                 data=CO2, method="REML")


#gratia draw() plot for CO2_modGS
draw(CO2_modGS)
```

```{r}
# setup prediction data
CO2_modGS_pred <- predict(CO2_modGS, se.fit=TRUE)
CO2 <- transform(CO2, 
                 modGS = CO2_modGS_pred$fit, 
                 modGS_se = CO2_modGS_pred$se.fit)

# plot the predictions
ggplot(data=CO2, aes(x=conc, y=uptake, color=Plant_uo, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(modGS-2*modGS_se),
                  ymax=exp(modGS+2*modGS_se)), alpha=0.25, color=NA) +
  geom_line(aes(y=exp(modGS))) +
  geom_point() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```

## Model 3: Global smoother plus group-level smoothers with different wiggliness
```{r}
CO2_modGI <- gam(log(uptake) ~ s(log(conc), k=5, m=2, bs="tp") +
                   s(log(conc), by= Plant_uo, k=5, m=1, bs="tp") +
                   s(Plant_uo, bs="re", k=12),
                 data=CO2, method="REML")

#plotting CO2_modGI 
draw(CO2_modGI, select = c(1,2,14,8,11,5), scales = "fixed")
```

```{r}
# setup prediction data
CO2_modGI_pred <- predict(CO2_modGI, se.fit=TRUE)
CO2 <- transform(CO2, 
                 modGI = CO2_modGI_pred$fit, 
                 modGI_se = CO2_modGI_pred$se.fit)

# plot the predictions
ggplot(data=CO2, aes(x=conc, y=uptake, color=Plant_uo, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(modGI-2*modGI_se),
                  ymax=exp(modGI+2*modGI_se)), alpha=0.25, color=NA) +
  geom_line(aes(y=exp(modGI))) +
  geom_point() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```

## AIC evaluation
```{r}
AIC_table <- AIC(CO2_modG,CO2_modGS, CO2_modGI) %>%
  rownames_to_column(var= "Model") %>%
  mutate(data_source = rep("CO2", 3)) %>%
  group_by(data_source) %>%
  mutate(deltaAIC = AIC - min(AIC)) %>%
  ungroup()%>%
  dplyr::select(-data_source) %>%
  mutate_at(.vars = vars(df,AIC, deltaAIC), 
            .funs = funs(round,.args = list(digits=0)))

AIC_table
```

# MPG dataset analysis
```{r}
# read in data
mpg.dat <- read.table("http://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data", 
                      header = FALSE)
colnames(mpg.dat) <- c("mpg", "cylinders", "displacement", "hp", "weight", "acceleration", 
                       "year", "origin", "car_name")
# data types
# str(mpg.dat)
# no missing values
# apply(mpg.dat, 2, function(x) any(is.na(x)))

# modify data types
mpg.dat$hp <- as.numeric(mpg.dat$hp)
mpg.dat$origin <- as.factor(mpg.dat$origin)

# remove car name from dataset
mpg.dat <- subset(mpg.dat, select = -c(car_name))

# shift year so that minimum starts at 0 not 1970
# re-lable origin from integer to character
mpg.dat <- mpg.dat %>%  
  mutate(
    year = year - min(year),
    origin = case_when(
      origin == 1 ~ "American",
      origin == 2 ~ "European",
      origin == 3 ~ "Japanese",
      TRUE ~ "other"
    )
  )

# convert origin into a factor variable
mpg.dat$origin <- factor(mpg.dat$origin, ordered = FALSE)

str(mpg.dat)
summary(mpg.dat)
```

## Selection of plots
We plot four of the covariates against `mpg`, color-coded by region of origin. Looking at `mpg` versus `year`, there seems to be a shared shape, even if the regions differ in the range of their responses. The acceleration plot, however, seems to suggest that the group-level smoothers need to be allowed differing wiggliness.

```{r}
mpg_dat_plot_yr <- ggplot(mpg.dat, 
                       aes(x=year, 
                           y=mpg, 
                           group=origin,
                           color=origin, 
                           lty=origin)) + 
  geom_point() +
  labs(title="MPG Data",
       x="Model Year since 1970", 
       y="mpg")

mpg_dat_plot_disp <- ggplot(mpg.dat,
                            aes(x=displacement,
                                y=mpg,
                                group=origin,
                                color=origin,
                                lty=origin)) +
  geom_point() +
  labs(title="MPG Data",
       x="Engine displacement in cubic inches", 
       y="mpg")

mpg_dat_plot_acc <- ggplot(mpg.dat,
                            aes(x=acceleration,
                                y=mpg,
                                group=origin,
                                color=origin,
                                lty=origin)) +
  geom_point() +
  labs(title="MPG Data",
       x="Time to accelerate from 0 to 60mph", 
       y="mpg")

mpg_dat_plot_hp <- ggplot(mpg.dat,
                            aes(x=hp,
                                y=mpg,
                                group=origin,
                                color=origin,
                                lty=origin)) +
  geom_point() +
  labs(title="MPG Data",
       x="Horsepower", 
       y="mpg")

plot_grid(mpg_dat_plot_yr,
          mpg_dat_plot_disp,
          mpg_dat_plot_acc,
          mpg_dat_plot_hp)
```

Displacement, cylinders, and weight are strongly correlated with each other, so I don't need to include all three in a model. I proceed with only displacement and the other covariates.

```{r warning=FALSE, message=FALSE}
GGally::ggpairs(subset(mpg.dat, select = c(mpg, cylinders, displacement, hp, weight, acceleration)))
```

## Training and Testing Data
I use even years for the training set, odd years for the test set.
```{r}
mpg.dat_train <- subset(mpg.dat, year %% 2 == 0)
mpg.dat_test <- subset(mpg.dat, year %% 2 == 1)
```

```{r}
#This function calculates the deviance of out-of-sample data,
#conditional on their mean predicted value from the model
get_deviance <- function(model, y_pred, y_obs, weights = NULL){
  stopifnot(length(y_obs)==length(y_pred))
  #We don't use the weights term in this paper, but it can be useful if
  #how well the model matters more for some sample points than others
  if(is.null(weights)) weights = rep(1, times= length(y_obs))
  #this uses the deviance residual function from the model family to
  #calculate deviances for individual points
  dev_residuals = model$family$dev.resids(y_obs, y_pred, weights)
  return(sum(dev_residuals))
}
```

## MPG Models 1, 2, and 3
```{r}
mpg_mod1 <- gam(mpg ~ s(displacement, bs="tp") +
                  s(hp, k=7, bs="tp") +
                  s(acceleration, bs="tp") +
                  s(year, k=5, bs="tp") +
                  s(origin, k=3, bs="re"),
                data = mpg.dat_train,
                family = "gaussian",
                method = "REML",
                drop.unused.levels = FALSE)

mpg_mod2 <- gam(mpg ~ s(displacement, bs="tp") +
                  s(displacement, origin, bs="fs", m=2) +
                  s(hp, bs="tp") +
                  s(hp, origin, bs="fs", m=2) +
                  s(acceleration, bs="tp") +
                  s(acceleration, origin, bs="fs", m=2) +
                  s(year, k=5, bs="tp") +
                  s(year, origin, k=3, bs="fs", m=2),
                data = mpg.dat_train,
                family = "gaussian",
                method = "REML",
                drop.unused.levels = FALSE)

mpg_mod3 <- gam(mpg ~ s(displacement, bs="tp", m=2) +
                  s(displacement, by=origin, k=7, bs="tp", m=1) +
                  s(hp, bs="tp", m=2) +
                  s(hp, by=origin, k=7, bs="tp", m=1) +
                  s(acceleration, bs="tp", m=2) +
                  s(acceleration, by=origin, k=7, bs="tp", m=1) +
                  s(year, k=5, bs="tp", m=2) +
                  s(year, by=origin, k=3, bs="tp", m=1) +
                  s(origin, k=3, bs="re"),
                data = mpg.dat_train,
                family = "gaussian",
                method = "REML",
                drop.unused.levels = FALSE)
```

## MPG Models 4 and 5
```{r}
mpg_mod4 <- gam(mpg ~ s(displacement, origin, bs="fs", m=2) +
                  s(hp, origin, bs="fs", m=2) +
                  s(acceleration, origin, bs="fs", m=2) +
                  s(year, origin, k=3, bs="fs", m=2),
                data = mpg.dat_train,
                family = "gaussian",
                method = "REML",
                drop.unused.levels = FALSE)

mpg_mod5 <- gam(mpg ~ s(displacement, by=origin, k=7, bs="tp", m=1) +
                  s(hp, by=origin, k=7, bs="tp", m=1) +
                  s(acceleration, by=origin, k=7, bs="tp", m=1) +
                  s(year, by=origin, k=3, bs="tp", m=1) +
                  s(origin, k=3, bs="re"),
                data = mpg.dat_train,
                family = "gaussian",
                method = "REML",
                drop.unused.levels = FALSE)
```

### Checking residuals and qqplots for model fit
We show QQ-plots with simulated confidence intervals. The deviance residuals from all the models look approximately normally distributed as we wanted. 
```{r}
par(mfrow=c(2,2))
gam.check(mpg_mod1)
gam.check(mpg_mod2)
gam.check(mpg_mod3)
gam.check(mpg_mod4)
gam.check(mpg_mod5)

plt1 <- qq_plot(mpg_mod5, method = "simulate") +
  labs(title =NULL, subtitle =NULL)
df <- data.frame(log_fitted = log(fitted(mpg_mod5)),
                 residuals  = resid(mpg_mod5, type = "deviance"))
plt2 <- ggplot(df, aes(x = log_fitted, y = residuals)) +
    geom_point() +
    labs(x = "Linear predictor", y = "Deviance residual")
plot_grid(plt1, plt2, ncol = 2, align = "hv", axis = "lrtb",labels=c("A","B"))
```

### Comparing out-of-sample predictions
Fit a null model with only an intercept for comparison:
```{r}
mpg_mod0 <- gam(mpg ~ s(origin, bs="re"),
                data = mpg.dat_train,
                family = "gaussian",
                method = "REML",
                drop.unused.levels = FALSE)
```

Get the out-of-sample predictions for all models:
```{r}
# We look at the correlation between fitted and observed values for all regions of origin:
mpg_test_summary <- mpg.dat_test %>%
  mutate(
    # out-of-sample predicted fits
    mod0 = predict(mpg_mod0, ., type="response"),
    mod1 = predict(mpg_mod1, ., type="response"),
    mod2 = predict(mpg_mod2, ., type="response"),
    mod3 = predict(mpg_mod3, ., type="response"),
    mod4 = predict(mpg_mod4, ., type="response"),
    mod5 = predict(mpg_mod5, ., type="response")
  ) %>%
  group_by(origin) %>%
  summarise(
    `Intercept only` = format(get_deviance(mpg_mod0, mod0, mpg),
                              scientific=FALSE,
                              digits=2),
    `Model 1` = format(get_deviance(mpg_mod1, mod1, mpg),
                       scientific=FALSE,
                       digits=2),
    `Model 2` = format(get_deviance(mpg_mod2, mod2, mpg),
                       scientific=FALSE,
                       digits=2),
    `Model 3` = format(get_deviance(mpg_mod3, mod3, mpg),
                       scientific=FALSE,
                       digits=2),
    `Model 4` = format(get_deviance(mpg_mod4, mod4, mpg),
                       scientific=FALSE,
                       digits=2),
    `Model 5` = format(get_deviance(mpg_mod5, mod5, mpg),
                       scientific=FALSE,
                       digits=2)
  )

mpg_test_summary
```

### AIC evaluation
```{r}
mpg_AIC_table <- AIC(mpg_mod1, mpg_mod2, mpg_mod3, mpg_mod4, mpg_mod5) %>%
  rownames_to_column(var= "Model") %>%
  mutate(data_source = rep("mpg.dat", 5)) %>%
  group_by(data_source) %>%
  mutate(deltaAIC = AIC - min(AIC)) %>%
  ungroup() %>%
  dplyr::select(-data_source) %>%
  mutate_at(.vars = vars(df, AIC, deltaAIC), 
            .funs = funs(round,.args = list(digits=0)))

mpg_AIC_table
```

